---
title: "DESeq2 Plots"
subtitle: "Prime-seq analysis"
author: "Data analysis: Paulo Jannig | Karolinska Institutet"
date: "`r paste('Last update:', format(Sys.time(), '%B %d, %Y'))`"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 3
    number_sections: true
    theme: spacelab
    highlight: default
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r}
print(setwd(dirname(rstudioapi::getSourceEditorContext()$path)))
```


```{r}
# define variables ==========================================
filename_prefix <- "" # leave empty if not necessary
species <- "mouse" # mouse, human, pig, rat
ensembl_version <- 102 # define ensembl release version

file_formats <- list(
  #"png",
  "pdf"
)

# load libraries
required_Packages_Install <- c("tidyverse",
                               "pheatmap",
                               "ggtext",
                               "RColorBrewer",
                               "scico",
                               "writexl",
                               "ggrepel",
                               "ggvenn",
                               "scales"
                               )


for(Package in required_Packages_Install){
  if(!require(Package,character.only = TRUE)) { 
    BiocManager::install(Package, dependencies=TRUE)
  }
  library(Package,character.only = TRUE)
}

# source custom functions
source("load_functions.R") # source custom functions
```

# Import data

## Normalized counts
```{r}
normalized_counts <- readRDS("../02.results/deseq2/deseq2_normalized_counts.Rds")
```

## Metadata
```{r}
metadata <- readRDS("../01.metadata/metadata.Rds")
colnames(metadata)
metadata$Sample
unique(metadata$Group)
```
## Contrasts
```{r}
group <- "STZ"
group_ref <- "Control"


contrast1 <- paste0(group,"_vs_",group_ref)
contrast1

```

```{r}
contrast1.df <- readRDS(paste0("../02.results/deseq2/deseq2_",
                               contrast1,"_lfcShrink.RDS"))

contrasts_list <- list(
  contrast1.df
)

# Assign names to the list elements
names(contrasts_list) <- c(contrast1)
```

```{r}
contrast1.df.sig <- contrast1.df %>% filter(padj<0.1)
#contrast2.df.sig <- contrast2.df %>% filter(padj<0.1)
```

## Heatmaps
```{r}
colnames(metadata)
unique(metadata$Group)
```

define color palette
```{r}
# Define colors
my_palette_heatmap <- list(
  Group = c(
    Control = "black",
    STZ = "#1F78B4"
  )
)

```

### all DEGs
for DEGs with padj < 0.05
```{r}
levels(metadata$Group)
colnames(metadata)
```


```{r}
# generate heatmap metadata (annotation_col)
heatmap_metadata <- metadata %>%
  arrange(Group) %>%
  dplyr::select(c(Sample, Group)) %>%
  column_to_rownames("Sample")

# retrieve genes of interest
#select DEGs for all contrasts (padj < 0.05)
heatmap_genes <- c()
for (contrast in contrasts_list) {
  genes <- contrast %>%
    filter(padj < 0.1) %>%
    pull(symbol)
  heatmap_genes <- union(heatmap_genes, genes)
  remove(genes, contrast)
}

# generate heatmap matrix
heatmap_counts <- normalized_counts %>%
  dplyr::filter(!!sym(symbol) %in% heatmap_genes)

heatmap_counts <- heatmap_counts %>%
  dplyr::arrange(desc(matrixStats::rowVars(as.matrix(
    heatmap_counts %>%
      dplyr::select(dplyr::matches(rownames(heatmap_metadata)))
  ))))
```

check for duplicated gene symbols
```{r}
duplicated_gene_symbols <- heatmap_counts %>%
  filter(duplicated(!!sym(symbol))) %>%
  distinct(!!sym(symbol))
duplicated_gene_symbols
```

```{r}
contrast1.df %>% filter(!!sym(symbol) %in% duplicated_gene_symbols)
```
```{r}
# Remove duplicates based on symbol, keeping the one with the highest baseMean
remove_genes <- contrast1.df %>% 
  filter(!!sym(symbol) %in% duplicated_gene_symbols) %>%
  arrange(!!sym(symbol), desc(baseMean)) %>%  # Sort by mgi_symbol and descending baseMean
  distinct(!!sym(symbol), .keep_all = TRUE) %>%   # Keep only the first occurrence of each mgi_symbol
  pull(ensembl_gene_id)

heatmap_counts <- heatmap_counts %>%
  filter(!ensembl_gene_id %in% remove_genes) %>%
  column_to_rownames(symbol) %>%
  dplyr::select(dplyr::matches(rownames(heatmap_metadata)))

```


Without gene labels
```{r}
heatmap <- pheatmap::pheatmap(heatmap_counts,
  cluster_rows = T,
  cluster_cols = F,
  scale = "row",
  clustering_distance_rows = "euclidean",
  #cutree_rows = 4,
  show_rownames = F, show_colnames = F,
  cellwidth = 3,
  #cellheight = 4,
  treeheight_row = 5,
  treeheight_col = 15,
  fontsize = 6,
  fontsize_row = 4,
  fontsize_col = 5,
  border_color = NA,
  gaps_col = c(5),
  annotation_colors = my_palette_heatmap,
  annotation_col = heatmap_metadata,
  color = colorRampPalette(
    rev(RColorBrewer::brewer.pal(11, "RdBu"))
  )(256)
)
```

Save plot
```{r}
filename <- "heatmap_significant_zcore_clustered"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 100,
    height = 30
  )
}
```

With gene labels
```{r}
heatmap <- pheatmap::pheatmap(heatmap_counts,
  cluster_rows = T,
  cluster_cols = F,
  scale = "row",
  clustering_distance_rows = "euclidean",
  #cutree_rows = 4,
  show_rownames = T, show_colnames = F,
  cellwidth = 3,
  cellheight = 4,
  treeheight_row = 5,
  treeheight_col = 15,
  fontsize = 6,
  fontsize_row = 4,
  fontsize_col = 5,
  border_color = NA,
  gaps_col = c(5),
  annotation_colors = my_palette_heatmap,
  annotation_col = heatmap_metadata,
  color = colorRampPalette(
    rev(RColorBrewer::brewer.pal(11, "RdBu"))
  )(256)
)
```

Save plot
```{r}
filename <- "heatmap_significant_zcore_clustered_labels"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 70,
    height = 200
  )
}
```

```{r echo=FALSE}
cat("Saved file :", paste0(
  "../03.figures/",
  filename_prefix, filename,
  ".pdf"
))
```

## MA Plots

define variables
```{r}
size_sig <- 0.5 # point size for the significant genes
size_ns <- 0.2 # point size for the non-significant genes

ylim_up <- 5 # upper limit for the y axis
ylim_down <- -2  # lower limit for the y-axis

# scico(10, palette = "navia")
# "#021326" "#06345E" "#14578A" "#27728E" "#388284" "#4A9279" "#65AB6C" "#98CB6F" "#D7E4A7" "#FCF3D8"
color_ns <- "#878787"      # color for non-significant values
color_low = "#67001F"      # color for p=0.00
color_high = "#4393C3"     # color for p=0.05

color_up <- "#B2182B"      # color for increased genes
color_down <- "#2166AC"    # color for decreased genes

significance_threshold <- 0.05

```


for selected contrast:
```{r}
group <- names(contrasts_list)[1] %>% str_extract(".*(?=_vs_)")
group_ref <- names(contrasts_list)[1] %>% str_extract("(?<=_vs_).*")
data <- contrasts_list[[1]]

plot_title <- paste0("**", group, " vs ", group_ref,"**")
plot_title

```


```{r}
ma <- ggplot(data, aes(
  x = log2(baseMean),
  y = log2FoldChange
)) +
  ggplot2::geom_point(
    data = filter(data, padj > significance_threshold | is.na(padj)),
    colour = color_ns,
    size = size_ns
  ) +
  ggplot2::geom_point(
    data = filter(data, padj < significance_threshold & log2FoldChange > 0) %>%
      arrange(-padj), colour = color_up, size = size_sig
  ) +
  ggplot2::geom_point(
    data = filter(data, padj < significance_threshold & log2FoldChange < 0) %>%
      arrange(-padj), colour = color_down, size = size_sig
  ) +
  # ggplot2::geom_point(
  #   data = filter(data, padj < significance_threshold) %>%
  #     arrange(-padj), aes(colour = padj), size = size_sig
  # ) +
  xlab("log<sub>2</sub> mean expression") +
  ylab("log<sub>2</sub> fold-change") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(plot_title) +
  gg_theme() +
  scale_color_gradient(
    low = color_low, high = color_high,
    limits = c(0, significance_threshold), oob = scales::squish
  ) +
  theme(legend.position = "right") +
  # geom_label_repel(
  #   data = filter(data, padj < significance_threshold & log2FoldChange > 0) %>% dplyr::slice(1:10),
  #   aes(label = !!sym(symbol)),
  #   size = 5 / ggplot2:::.pt,
  #   force = 2,
  #   label.padding = 0.1,
  #   # face
  #   direction = "both",
  #   max.overlaps = 10,
  #   #nudge_x = -2,
  #   #nudge_y = 1,
  #   min.segment.length = 0,
  #   segment.size = 0.25
  # )  +
  # geom_label_repel(
  #   data = filter(data, padj < significance_threshold & log2FoldChange < 0) %>% dplyr::slice(1:10),
  #   aes(label = !!sym(symbol)),
  #   size = 5 / ggplot2:::.pt,
  #   force = 4,
  #   label.padding = 0.1,
  #   # face
  #   direction = "both",
  #   max.overlaps = 10,
  #  # nudge_x = -2,
  #   #nudge_y = -1,
  #   min.segment.length = 0,
  #   seed = 1234,
  #   segment.size = 0.25
  # ) +
  coord_cartesian(ylim = c(ylim_down, ylim_up), expand = T) +
  #geom_hline(yintercept = 1, linetype = "dotted") +
  #geom_hline(yintercept = -1, linetype = "dotted") +
  annotate("text",
    x = max(log2(data$baseMean)), 
    #y = max(data$log2FoldChange),
    y = ylim_up,
    label = paste0(
      #group,": ",
      "Increased: ",
      nrow(filter(data, padj < significance_threshold & log2FoldChange > 0))
    ),
    hjust = 1, vjust = 1, size = 5 / ggplot2:::.pt
  ) +
  annotate("text",
    x = max(log2(data$baseMean)),
    #y = min(data$log2FoldChange),
    y = ylim_down,
    label = paste0(
      #group_ref,
      "Decreased: ",
      nrow(filter(data, padj < significance_threshold & log2FoldChange < 0))
    ),
    hjust = 1, vjust = 0, size = 5 / ggplot2:::.pt
  )
ma
```
Save plot
```{r}

filename <- paste0(group, "_vs_", group_ref)
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      "MA_plot_",
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 30,
    plot_height = 30,
    width = 60,
    height = 50
  )
}
```


# R session info
```{r}
utils:::print.sessionInfo(sessionInfo()[-8])
```